The Design, Semantics, and Implementation of \CEU: a Synchronous Reactive Language based on Esterel

### 2015-10-05 paper.tex
Pequenos ajustes.

\begin{abstract}
\CEU is a reactive language based on Esterel that targets constrained embedded 
platforms and ensures safe concurrency by handling threats at compile time, 
rather than at runtime.
%
Based on the synchronous programming model, our design allows for a simple 
reasoning about concurrency that enables compile-time analysis resulting in 
deterministic and memory-safe programs.
%
% TODO: some uses?
%
In this work, we discuss the design of \CEU and propose a formal semantics 
focusing on its particular control mechanisms, such as parallel compositions, 
finalization, and stack-based internal events.
%
We also present an implementation with two back ends:
one aiming for resource efficiency and interoperability with \C,
and another based on a virtual machine that allows remote reprogramming.
\end{abstract}

@@

In previous work~\cite{ceu.sensys13,ceu.terra}, we employed \CEU in the context 
of wireless sensor networks, developing a number of applications, protocols, 
and drivers.
%
We evaluated the expressiveness of \CEU in comparison to event-driven code in 
\C and attested a reduction in source code size (around 25\%) with a small 
increase in memory usage (around 5--10\% for \emph{text} and 
\emph{data})~\cite{ceu.sensys13}.
%
Considering the \VM back end, a simple application that blinks three LEDs 
periodically occupies less than 100 bytes and can be completely transmitted in 
4 radio messages.
In contrast, a \C-compiled version occupies more than 2 Kbytes (as it includes 
all \emph{I/O} functionality preloaded with the \VM)~\cite{ceu.terra}.

@@

In the sections that follow, we discuss the main differences between \CEU and 
Esterel~\cite{ceu.sensys13}:
queue-based external events and stack-based internal events 
(Section~\ref{sec.ceu.ints}),
shared-memory concurrency and determinism (Section~\ref{sec.ceu.det}),
safe abortion with finalization (Section~\ref{sec.ceu.abrt}),
and first-class timers (Section~\ref{sec.ceu.timers}).

Regarding the similarities, Figure~\ref{lst.abro} shows side-by-side the 
implementations in Esterel and \CEU for the following control 
specification~\cite{esterel.primer}:
%
\emph{``Emit an output O as soon as two inputs A and B have occurred.
Reset this behavior each time the input R occurs''.}
%
The first phrase of the specification, awaiting and emitting the events, is 
translated almost identically in the two languages (lines 4--9, in both 
implementations), given that Esterel's `$\|$' and \CEU's \code{par/and} 
constructs are equivalent.
%
For the second phrase, the reset behavior, the Esterel version uses a 
\code{abort-when} (lines 3--10), which serves the same purpose of \CEU's 
\code{par/or} (lines 3--12):
the occurrence of event \code{R} aborts the awaiting statements in parallel and 
restarts the \code{loop}.

@@

\CEU, following the Esterel mindset, has a strong imperative flavor, with 
explicit control flow through sequences, loops, parallels, and also 
assignments.
Being designed for control-intensive applications, it provides support for 
concurrent lines of execution and broadcast communication through events.
%
In addition, \CEU also employs the synchronous model, in which programs advance 
in a sequence of discrete reactions to external events.
Internal computations within a reaction (e.g. expressions, assignments, and 
system calls) are considered to take no time in accordance with the synchronous 
hypothesis~\cite{rp.hypothesis}.
The \code{await} statements are the only ones that halt a running reaction and 
allow a program to advance in this notion of time.
%
To ensure that reactions run in bounded time and programs always progress, 
loops are statically required to contain at least one \code{await} statement in 
all possible paths~\cite{ceu.sensys13,esterel.primer}.
%
\CEU shares the same limitations with Esterel and synchronous languages in 
general:
computations that run in unbounded time (e.g., cryptography, image processing) 
do not fit the zero-delay hypothesis~\cite{rp.hypothesis}, and cannot be 
elegantly implemented.

@@

\begin{figure}
\begin{minipage}[t]{0.43\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em,mathescape=true]
// ESTEREL
input A;    // external
signal B;   // internal
[[
    await A;
    emit B;
    call f();
$\|$
    await B;
    call g();
]]
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.53\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.1em]
// CEU
input void A;  // external (in uppercase)
event void b;  // internal (in lowercase)
par/and do
    await A;
    emit b;
    _f();
with
    await b;
    _g();
end
\end{lstlisting}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Internal signals (events) in Esterel and \CEU. \newline
\label{lst.prints}
}
\end{figure}

@@

Another distinction regarding event handling in comparison to \CEU is that 
Esterel supports same-cycle bi-directional 
communication~\cite{esterel.compiling}, i.e., two threads can react to one 
another during the same cycle due to mutual signal dependency.
%
\CEU imposes a restriction that an \code{await} to an internal event is only 
valid for the next reaction, i.e., if an \code{await} and \code{emit} occur 
simultaneously in parallel trails, the \code{await} does not awake.
%
These \emph{delayed awaits} avoid corner cases of instantaneous termination and 
re-execution of statements in the same reaction inside loops (known as 
\emph{schizophrenic statements}~\cite{esterel.schizo,esterel.schizo2}).
%
The example that follows relies on this restriction to avoid infinite 
execution:

\begin{lstlisting}
event void e,f;
loop do
    par/or do
        await e;
    with
        emit e;     // without the restriction, it awakes 1s trail and restarts the loop
        await f;
    end
end
\end{lstlisting}

@@

In atypical scenarios requiring immediate awake, delayed awaits can be 
circumvented by manually copying or transforming the code to execute on awake.
%
From our experience, in some cases we need to execute a block of code 
periodically from internal event requests, \emph{including the current 
reaction}, as illustrated in the left of Figure~\ref{lst.delay}.
In this case, moving the \code{await} to the end of the loop (line 10) makes 
the periodic code to also execute immediately (line 9), and then in reactions 
to each \code{emit} request (line 5).
If the periodic \code{emit} depends on a condition, as illustrated in the right 
of Figure~\ref{lst.delay}, the code becomes more intricate because we need a 
state variable (line 2) and to copy the condition test to the periodic code 
(line 13).
%
On the one hand, we transfer the burden of dealing with these corner cases to 
the programmer.
On the other hand, we simplify the semantics of the language and eliminate the 
need for analysis to deal with shizophrenic statements.

@@

Concurrency in \CEU is characterized when two or more trail segments in 
parallel execute during the same reaction chain.
A trail segment is a sequence of statements followed by an \code{await} (or 
termination).
%
In the program in the left of Figure~\ref{lst.shared}, the two assignments to 
\code{x} can never run concurrently, because each trail segment reacts to a 
different input event and, according to the semantics of \CEU, cannot occur 
simultaneously.
%
However, the program in the right is non-deterministic, because the two 
assignments to \code{y} occur in the same reaction to input \code{A}.

\begin{figure}
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}
input void A, B;
var int x = 1;
par/and do
    await A;
    x = x + 1;
with
    await B;
    x = x * 2;
end
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}
input void A;
var int y = 1;
par/and do
    await A;
    y = y + 1;
with
    await A;
    y = y * 2;
end
\end{lstlisting}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Shared-memory concurrency in \CEU:
the code in the left is safe because the trails access \code{x} atomically in 
different reactions;
the code in the right is unsafe because both trails access \code{y} in the same 
reaction.
\label{lst.shared}
}
\end{figure}

\CEU performs a temporal analysis at compile time and detects concurrent 
accesses to shared variables, as follows~\cite{ceu.sensys13}:
\emph{if a variable is written in a trail segment, then a concurrent trail 
segment cannot read or write to that variable, nor dereference a pointer of 
that variable type.}
An analogous policy is applied for pointers \emph{vs} variables and pointers 
\emph{vs} pointers, as well as for system calls with side effects (e.g., 
\code{printf}).
%The algorithm for the analysis holds the set of all events in preceding 
%\code{await} statements for each variable access.
%Then, the sets for all accesses in parallel trails are compared to assert that 
%no events are shared among them.
%Otherwise the compiler warns about the suspicious accesses.

Regardless of the temporal analysis of \CEU, when multiple trails are active 
during the same reaction, they are scheduled in the order they appear in the 
program source code.
%
Therefore, even though the program in the right of Figure~\ref{lst.shared} is 
suspicious, the assignments to \code{y} are both atomic and deterministic, 
i.e., after the reaction to \code{A} terminates, the value of \code{y} is $4$ 
(\code{(1+1)*2}).
%
On the one hand, enforcing an execution order for concurrent operations may 
seen arbitrary and also precludes true parallelism.
On the other hand, it provides a priority scheme for trails, and makes 
shared-memory concurrency more tractable.
%
For constrained embedded development, we believe that deterministic 
shared-memory concurrency is beneficial, given the extensive use of memory 
mapped ports for I/O and the lack of hardware support for real parallelism.
Other synchronous embedded languages, such as \emph{SOL}~\cite{wsn.sol} and 
\emph{PRET-C}\cite{rp.pretc}, made a similar design choice.
%For Esterel, however, is also used in hardware design~\cite{rp.twelve} where 
%parallelism is inherent.

@@

To illustrate threats related to abortion, consider the unsafe example, which 
does not compile, in the left of Figure~\ref{lst.fin}, describing the state 
machine of a data collection protocol for sensor 
networks~\cite{wsn.ctp,ceu.sensys13}.
%
The input events \code{STOP}, \code{RETRANSMIT}, and \code{SENDACK} (line 1) 
represent the external interface of the protocol.
%
The protocol has to transmit a packet every minute, unless it receives a 
\code{RETRANSMIT} request to immediately re-transmit it, or a \code{STOP} 
request to terminate.
%
The protocol is composed of two main trails:
one simply monitors the stopping event (line 3);
the other periodically transmits the packet (lines 5--18).
%
The periodic transmission is a loop that starts two other trails (lines 6--17):
one handles the immediate retransmission request (line 7);
the other transmits the packet%
\footnote{
The underline prefix marks (e.g., \code{\_send\_enqueue}) make interactions 
with external \C libraries explicit and are required in \CEU.
}
and waits for a confirmation (lines 9--16).
%
%
The actual transmission (lines 12--15) is enclosed with a \code{par/and} that 
takes at least one minute before looping, in accordance with the specification.
Note that the call to \code{\_send\_enqueue} is \emph{asynchronous}, handing to 
the radio driver a pointer to the lexically-scoped packet.
The driver makes the transmission in the background, holding the packet until 
it signals the application with the \code{SENDACK} to acknowledge completion.
%
At any time, the client may request a retransmission (line 7), which terminates 
the \code{par/or} (line 6), aborts the ongoing transmission (line 14, if not 
idle), and restarts the loop (line 5).
%
The client may also request to stop the whole protocol at any time (line 3).
%
Therefore, if the sending trail is aborted by the \code{STOP} or 
\code{RETRANSMIT} requests, the packet buffer goes out of scope, leaving behind 
a \emph{dangling pointer} in the radio driver, which will possibly transmit 
corrupted data.
%
%\subsubsection{Finalization}
%\label{sec.ceu.fin}

The unsafe example in the left of Figure~\ref{lst.fin} does not compile because 
\CEU tracks the interaction of \code{par/or} compositions with local variables 
and stateful \C functions (e.g., device drivers) in order to preserve safe 
abortion of trails~\cite{ceu.sensys13,ceu.mod15}.
%
%Finalization clauses are fundamental to preserve the orthogonality of 
%\code{par/or} compositions in SSRP.
%
In fact, \CEU enforces the programmer to write a \emph{finalization} clause to 
accompany the stateful \C call.
The code in the right of Figure~\ref{lst.fin} properly cancels the packet 
transmission when the block of \code{buffer} goes out of scope, i.e., the 
finalization clause (after the \code{with}) executes automatically on external 
abortion.%
\footnote{
Note that the compiler only enforces the programmer to write the finalization 
clause, but cannot check if it actually handles the resource properly.
}
%
%Note also that \CEU environments rely on \C libraries that only provide 
%asynchronous I/O and non-blocking functions~\cite{ceu.sensys13}.

@@

The \code{await} statement of \CEU supports wall-clock time and handles deltas 
automatically~\cite{ceu.sensys13}, resulting in more robust applications.
For the example in the left of Figure~\ref{lst.timers}, suppose that after the 
first \code{await} request, the underlying system gets busy and takes 15ms to 
check for expiring awaits.
The \CEU scheduler will notice that the \code{await 10ms} has not only already 
expired, but is delayed with \code{delta=5ms}.
Then, the awaiting trail awakes, sets \code{v=1}, and invokes \code{await 1ms}.
As the current delta is higher than the requested timeout (i.e. $5ms > 1ms$), 
the trail is rescheduled for execution, now with \code{delta=4ms}.

>>>>

\begin{abstract}
\CEU is a reactive language based on Esterel that targets constrained embedded 
platforms and ensures safe concurrency by handling threats at compile time, 
rather than at runtime.
%
Based on the synchronous programming model, our design allows for a simple 
reasoning about concurrency that enables compile-time analysis and results in 
deterministic and memory-safe programs.
%
% TODO: some uses?
%
In this work, we discuss the design of \CEU and propose a formal semantics 
focusing on its particular control mechanisms, such as parallel compositions, 
finalization, and stack-based internal events.
%
We also present an implementation with two back ends:
one aiming for resource efficiency and interoperability with \C,
and another based on a virtual machine that allows remote reprogramming.
\end{abstract}

@@

In previous work~\cite{ceu.sensys13,ceu.terra}, we employed \CEU in the context 
of wireless sensor networks, developing a number of applications, protocols, 
and drivers.
%
We evaluated the expressiveness of \CEU in comparison to event-driven code in 
\C and attested a reduction in source code size (around 25\%) with a small 
increase in memory usage (around 5--10\% for \emph{text} and 
\emph{data})~\cite{ceu.sensys13}.
%
Considering the \VM back end, a simple application that blinks three LEDs 
periodically occupies less than 100 bytes and can be completely transmitted in 
4 radio messages, given that \emph{I/O} components are preloaded in the remote
devices.
In contrast, a \C-compiled version occupies more than 2 Kbytes, which includes 
all \emph{I/O} functionality~\cite{ceu.terra}.

@@

In the sections that follow, we discuss the main differences between \CEU and 
Esterel:
queue-based external events and stack-based internal events 
(Section~\ref{sec.ceu.ints}),
shared-memory concurrency and determinism (Section~\ref{sec.ceu.det}),
safe abortion with finalization (Section~\ref{sec.ceu.abrt}),
and first-class timers (Section~\ref{sec.ceu.timers}).

Regarding the similarities, Figure~\ref{lst.abro} shows side-by-side the 
implementations in Esterel and \CEU for the following control 
specification~\cite{esterel.primer}:
%
\emph{``Emit an output O as soon as two inputs A and B have occurred.
Reset this behavior each time the input R occurs''.}
%
The first phrase of the specification, awaiting and emitting the events, is 
translated almost identically in the two languages (lines 4--9, in both 
implementations), given that Esterel's `$\|$' and \CEU's \code{par/and} 
constructs are equivalent.
%
For the second phrase, the reset behavior, the Esterel version uses a 
\code{abort-when} (lines 3--10), which serves the same purpose of \CEU's 
\code{par/or} (lines 3--12):
the occurrence of event \code{R} aborts the awaiting statements in parallel and 
restarts the enclosing \code{loop}.

@@

\CEU, following the Esterel mindset, has a strong imperative flavor, with 
explicit control flow through sequences, loops, parallels, and also 
assignments.
Being designed for control-intensive applications, it provides support for 
concurrent lines of execution and broadcast communication through events.
%
In addition, \CEU also employs the synchronous model, in which programs advance 
in a sequence of discrete reactions to external events.
Internal computations within a reaction (e.g. expressions, assignments, and 
system calls) are considered to take no time in accordance with the synchronous 
hypothesis~\cite{rp.hypothesis}.
The \code{await} statements are the only ones that halt a running reaction and 
allow a program to advance in this notion of time.
%
To ensure that reactions run in bounded time and programs always progress, 
loops are statically required to contain at least one \code{await} statement in 
all possible paths~\cite{ceu.sensys13,esterel.primer}.
%
\CEU shares the same limitations with Esterel and synchronous languages in 
general:
computations that run in unbounded time (e.g., cryptography, image processing) 
do not fit the zero-delay hypothesis, and cannot be elegantly implemented.

@@

\begin{figure}
\begin{minipage}[t]{0.43\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em,mathescape=true]
// ESTEREL
input A;    // external
signal B;   // internal
[[
    await A;
    emit B;
    call f();
$\|$
    await B;
    call g();
]]
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.53\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.1em]
// CEU
input void A;  // external (in uppercase)
event void b;  // internal (in lowercase)
par/and do
    await A;
    emit b;
    _f();
with
    await b;
    _g();
end
\end{lstlisting}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Internal signals (events) in Esterel and \CEU: similar syntax, but 
different semantics.
\label{lst.prints}
}
\end{figure}

@@

Another distinction regarding event handling in comparison to \CEU is that 
Esterel supports same-cycle bi-directional 
communication~\cite{esterel.compiling}, i.e., two threads can react to one 
another during the same cycle due to mutual signal dependency.
%
\CEU imposes a restriction that an \code{await} is only valid for the next 
reaction, i.e., if an \code{await} and \code{emit} occur simultaneously in 
parallel trails, the \code{await} does not awake.
%
These \emph{delayed awaits} avoid corner cases of instantaneous termination and 
re-execution of statements in the same reaction inside loops (known as 
\emph{schizophrenic statements}~\cite{esterel.schizo,esterel.schizo2}).
%
The example that follows relies on this restriction to avoid infinite 
execution:

\begin{lstlisting}
event void e,f;
loop do
    par/or do
        await e;
    with
        emit e; // w/o the restriction, awakes 1st trail and restarts the loop instantaneously
        await f;
    end
end
\end{lstlisting}

@@

In atypical scenarios requiring immediate awake, delayed awaits can be 
circumvented by manually copying or transforming the code to execute on awake.
%
From our experience, in some cases we need to execute a block of code 
periodically from internal event requests, \emph{including the current 
reaction}, as illustrated in the left of Figure~\ref{lst.delay}.
In this case, moving the \code{await} to the end of the loop (line 10) makes 
the periodic code to also execute immediately (line 9), and then in reactions 
to each \code{emit} request (line 5).
If the periodic \code{emit} depends on a condition, as illustrated in the right 
of Figure~\ref{lst.delay}, the code becomes more intricate because we need a 
state variable (line 2) and also to copy the condition test to the periodic 
code (line 13).
%
On the one hand, we transfer the burden of dealing with these corner cases to 
the programmer.
On the other hand, we simplify the semantics of the language and eliminate the 
need for analysis to deal with schizophrenic statements.

@@

Concurrency in \CEU is characterized when two or more trail segments in 
parallel execute during the same reaction chain.
A trail segment is a sequence of statements followed by an \code{await} (or 
termination).
%
In the program in the left of Figure~\ref{lst.shared}, the two assignments to 
\code{x} (lines 5,8) can never run concurrently, because each trail segment 
reacts to a different input event (lines 4,7) and, according to the semantics 
of \CEU, cannot occur simultaneously.
%
However, the program in the right is non-deterministic, because the two 
assignments to \code{y} (lines 5,8) occur in the same reaction to input 
\code{A} (lines 4,7).

\begin{figure}
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
input void A, B;
var int x = 1;
par/and do
    await A;
    x = x + 1;
with
    await B;
    x = x * 2;
end
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
input void A;
var int y = 1;
par/and do
    await A;
    y = y + 1;
with
    await A;
    y = y * 2;
end
\end{lstlisting}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Shared-memory concurrency in \CEU:
the code in the left is safe because the trails access \code{x} atomically in 
different reactions;
the code in the right is unsafe because both trails access \code{y} in the same 
reaction.
\label{lst.shared}
}
\end{figure}

\CEU performs a temporal analysis at compile time and detects concurrent 
accesses to shared variables, as follows:
\emph{if a variable is written in a trail segment, then a concurrent trail 
segment cannot read or write to that variable, nor dereference a pointer of 
that variable type.}
An analogous policy is applied for pointers \emph{vs} variables and pointers 
\emph{vs} pointers, as well as for system calls with side effects (e.g., 
\code{printf}).
%The algorithm for the analysis holds the set of all events in preceding 
%\code{await} statements for each variable access.
%Then, the sets for all accesses in parallel trails are compared to assert that 
%no events are shared among them.
%Otherwise the compiler warns about the suspicious accesses.

Regardless of the temporal analysis of \CEU, when multiple trails are active 
during the same reaction, they are scheduled in the order they appear in the 
program source code.
%
Therefore, even though the program in the right of Figure~\ref{lst.shared} is 
suspicious, the assignments to \code{y} are both atomic and deterministic, 
i.e., after the reaction to \code{A} terminates, the value of \code{y} is $4$ 
(\code{(1+1)*2}).
%
On the one hand, enforcing an execution order for concurrent operations may 
seen arbitrary and also precludes true parallelism.
On the other hand, it provides a priority scheme for trails, and makes 
shared-memory concurrency more tractable.
%
For constrained embedded development, we believe that deterministic 
shared-memory concurrency is beneficial, given the extensive use of memory 
mapped ports for \emph{I/O} and the lack of hardware support for real 
parallelism.
Other synchronous embedded languages, such as \emph{SOL}~\cite{wsn.sol} and 
\emph{PRET-C}~\cite{rp.pretc}, made a similar design choice.
%For Esterel, however, is also used in hardware design~\cite{rp.twelve} where 
%parallelism is inherent.

@@

To illustrate threats related to abortion, consider the unsafe example in the 
left of Figure~\ref{lst.fin}, which does not compile.
It describes the state machine of a data collection protocol for sensor 
networks~\cite{wsn.ctp,ceu.sensys13}.
%
The input events \code{STOP}, \code{RETRANSMIT}, and \code{SENDACK} (line 1) 
represent the external interface of the protocol.
%
The protocol has to transmit a packet every minute, unless it receives a 
\code{RETRANSMIT} request to immediately re-transmit it, or a \code{STOP} 
request to terminate.
%
The protocol is composed of two main trails:
one simply monitors the stopping event (line 3);
the other periodically transmits the packet (lines 5--18).
%
The periodic transmission is a loop that starts two other trails (lines 6--17):
one handles the immediate retransmission request (line 7);
the other transmits the packet and waits for a confirmation (lines 9--16).
%
%
The actual transmission (lines 12--15) is enclosed with a \code{par/and} that 
takes at least one minute before looping, in accordance with the specification.
Note that the call to \code{\_send\_enqueue}%
\footnote{
The underline prefix marks (e.g., \code{\_send\_enqueue}) make interactions 
with external \C libraries explicit and are required in \CEU.
}
(line 14) is \emph{asynchronous}, handing to the radio driver a pointer to the 
lexically-scoped packet (line 12).
The driver makes the transmission in the background, holding the packet until 
it signals the application with the \code{SENDACK} to acknowledge completion 
(line 15).
%
At any time, the client may request a retransmission (line 7), which terminates 
the \code{par/or} (line 6), aborts the ongoing transmission (line 14, if not 
idle), and restarts the loop (line 5).
%
The client may also request to stop the whole protocol at any time (line 3).
%
Therefore, if the sending trail is aborted by the \code{STOP} or 
\code{RETRANSMIT} requests, the packet buffer goes out of scope, leaving behind 
a \emph{dangling pointer} in the radio driver, which will possibly transmit 
corrupted data.
%
%\subsubsection{Finalization}
%\label{sec.ceu.fin}

The unsafe example in the left of Figure~\ref{lst.fin} does not compile because 
\CEU tracks the interaction of \code{par/or} compositions with local variables 
and stateful \C functions (e.g., device drivers) in order to preserve safe 
abortion of trails.
%
%Finalization clauses are fundamental to preserve the orthogonality of 
%\code{par/or} compositions in SSRP.
%
In fact, \CEU enforces the programmer to write a \emph{finalization} clause to 
accompany the stateful \C call.
The code in the right of Figure~\ref{lst.fin} properly cancels the packet 
transmission when the block of \code{buffer} goes out of scope, i.e., the 
finalization clause (after the \code{with}) executes automatically on external 
abortion.%
\footnote{
Note that the compiler only enforces the programmer to write the finalization 
clause, but cannot check if it actually handles the resource properly.
}
%
%Note also that \CEU environments rely on \C libraries that only provide 
%asynchronous I/O and non-blocking functions~\cite{ceu.sensys13}.

@@

The \code{await} statement of \CEU supports wall-clock time and handles deltas 
automatically, resulting in more robust applications.
For the example in the left of Figure~\ref{lst.timers}, suppose that after the 
first \code{await} request, the underlying system gets busy and takes 15ms to 
check for expiring awaits.
The \CEU scheduler will notice that the \code{await 10ms} has not only already 
expired, but is delayed with \code{delta=5ms}.
Then, the awaiting trail awakes, sets \code{v=1}, and invokes \code{await 1ms}.
As the current delta is higher than the requested timeout (i.e. $5ms > 1ms$), 
the trail is rescheduled for execution, now with \code{delta=4ms}.
